---
title: "SMPA - Grouping Assignment"
author: "Bryn McLaughlin"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    self-contained: true
---

```{r}

#load libraries
library(tidyverse)
library(lubridate)
#load data
payments <- readRDS("data/trump_campaign_payments.rds")


```

## Introduction

**What data will you be working with?**

A dataset of disbursements (payments and purchases) by the Donald J. Trump for President campaign committee during the 2020 election cycle. The included payments are from 2017 and 2018.

This data and examples were developed as part of the NYT Open training, which we've adapted from here. Note that the dataset has been alterted for instructional purposes, so doesn't represent the original FEC data in its entirety.

In the environment tab in the upper right, click on the `payments` dataset to open the RStudio viewer to examine the data. You can also run the chunk below as well to bring up a slice of the data.

```{r}

head(payments)


```

## Before you begin

Take a moment to look at the various categories included for each disbursement.

-   Notice that there are several date-related fields:

    -   FEC election cycle

    -   report year

    -   election type

    -   disbursement date

-   There are a few fields about the recipient. Along with city and state, you have:

    -   the recipient name

    -   the payee's first and last name (an individual who was paid for the "recipient") and

    -   the entity type (individual v. organization)

## Initial Narrative Questions

Without doing any analysis yet, what are three questions you might try to answer using this dataset? Write your narrative answers in the spaces below.

1.  What state received the most funds from the Committee?

2.  What organization recieved the most individual disburstments from the Committee?

3.  During what month did the Committee spend the most money?

## Coding Assignment Questions

**Please write your R tidyverse code to help answer the questions below in the provided R chunks. You can make more than one chunk if you prefer.**

**Also, describe what you're doing either in text out "in the open" above the chunk, or in comments (prefaced by #) in the R chunk itself.**

**You'll need to use grouping, filtering, or a combination of both to find the answers. Remember that R is case-sensitive when you refer to any text values within the columns**

<br>

How much was spent in the 2017 vs. 2018 reporting year?

```{r}

payments %>%
  select (report_year, disbursement_amount) %>% 
  filter(report_year == 2017 |2018) %>% 
  group_by(report_year) %>% 
  summarise(total = sum(disbursement_amount))
  
# I selected the reporting year and amount, filtered to only 2017 and 2018, grouped by report year, and made the sum of disbursements a new aggregate total.

```

Tally the total overall disbursement amounts for each month (distinguishing between say, Oct. 2018 v. Oct. 2019). In which month was spending the highest?

```{r}

payments  %>%
  select (report_year,disbursement_month, disbursement_amount) %>% 
  group_by(report_year,disbursement_month) %>% 
  summarise(total = sum(disbursement_amount))

#selected year, month, and amount, grouped by year and month, turned sum of each months spending into a new aggregate variable.

```

Which recipient (recipient_name) was paid the most in the 2020 election cycle?

```{r}

 payments  %>%
  filter(election_cycle == 2020) %>% 
  group_by(recipient_name) %>% 
  summarise(total = sum(disbursement_amount)) %>% 
  arrange(desc(total))

#Filtered to only election cycle 2020, group by recipient name, summed the disbursement amount, then arranged the aggregate 'total' variable in descending order. The highest paid recipient was AMERICAN MADE MEDIA CONSULTANTS, LLC.

```

Look at what this recipient was paid for. What service or good (see the "disbursement_description") were they paid the most for?

```{r}

payments  %>%
  filter(election_cycle == 2020, recipient_name == "AMERICAN MADE MEDIA CONSULTANTS, LLC") %>% 
  group_by(disbursement_description) %>% 
  summarise(count=n())

# "AMERICAN MADE MEDIA CONSULTANTS, LLC" was paid most for digital consulting services. I filtered to the year 2020 and only the recipient, then grouped by disbursement description and summarized the count in a new aggregate variable.


```

What's the most common category overall (see "disbursement_description") the campaign spent money on?

```{r}

payments  %>%
  select(disbursement_description) %>% 
  group_by(disbursement_description) %>% 
  summarise(count=n()) %>% 
  arrange(desc(count))
  
#The most common category overall was merchant fees. I selected the description category, grouped by entry, the summarized and arranged a new aggregate variable in descending order.

```

Who were the 5 biggest recipients of campaign spending on "LEGAL CONSULTING"..?

```{r}

payments  %>%
  select(recipient_name, disbursement_description, disbursement_amount) %>% 
  filter(disbursement_description == "LEGAL CONSULTING") %>% 
  group_by(recipient_name) %>% 
  summarise(count=n(), total = sum(disbursement_amount)) %>% 
  arrange(desc(total))
  
#The 5 biggest recipients of campaign spending on legal consulting are JONES DAY, LAROCCA HORNIK ROSEN GREENBERG & BLAHA LLP, KASOWITZ, BENSON, TORRESS & FRIEDMAN, THE LAW OFFICES OF ALAN S. FUTERFAS, and MCDERMOTT WILL & EMERY. I selected by recipient, disbursement description and amount, filtered by legal consulting, grouped by recipeint name, summarized count and created a new total aggregate variable for the disbursement amount, then arranged total by descending order to find the top 5 disbursements.

```


Who are the top 5 individuals (vs. organization) were paid by the campaign committee?

```{r}

payments %>%
  select (recipient_name, entity_type, disbursement_amount) %>% 
  filter(entity_type == "IND") %>% 
  group_by(recipient_name, entity_type) %>% 
  summarise(count=n(), total = sum(disbursement_amount)) %>% 
  arrange(desc(total))

#The top five individuals paid the most are GLASSNER, MICHAEL; DOLLMAN, SEAN; CENTINELLO, DARREN; CARR, CHRISTOPHER; and PENCE, JOHN. I selected recipeints, entity type, and disbursement amount, filtered for individuals, grouped by recipient name and entity type, summarized the count and total disbursement amount, then arranged the total in descending order.



```


For the indvidual who was paid the most, which payment category (disbursement_description) was the top one for him or her? And how much money were they paid in that category?

```{r}

payments %>%
  select (recipient_name, disbursement_description, disbursement_amount) %>% 
  filter(recipient_name == "GLASSNER, MICHAEL") %>% 
  group_by(disbursement_description) %>% 
  summarise(count=n(), total = sum(disbursement_amount)) %>% 
  arrange(desc(total))

#The top payment category for "GLASSNER, MICHAEL" was PAYROLL, where he was paid $499000. I selected recipeint name, disbursemnt description and amount, then filtered for the individual. Next I grouped by description, summarized count and disbursement sum in the aggregate variable "total", then arranged by descending total.
```
